Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada

### 2024-04-19 paper/paper.qmd
finalize paper, update readme, update reference, fix grammar

# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("here")
# install.packages("mapdata")
# install.packages("ggplot2")
# install.packages("haven")
# install.packages("kableExtra")
# install.packages('tinytex')
# tinytex::install_tinytex()
# to uninstall TinyTeX, run tinytex::uninstall_tinytex()

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
library(tinytex)
library(modelsummary)
```

@@

In this study, our primary focus is to estimate the effect of education on hourly wages in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.

The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. @sec-model then delves into the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation. After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data. @sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.

@@

This study utilizes a dataset sourced from the Open Government Portal of Canada [], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce. The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.

@@

The data was processed and cleaned using R[@citeR], a powerful statistical programming language. The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends. Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research. This step was crucial to maintaining the clarity and relevance of the analysis. For key operations, please refer to the [@sec-appendix].

@@

```{r}
#| label: tbl-display-clean-data
#| tbl-cap: First Ten Rows of Wage Rates Spanning from 2000 to 2019 For Cleaned Data
#| echo: false

@@

```{r fig.width=7, fig.height=3}
#| label: fig-display-clean-data
#| fig-cap: Visualization of Wage Rate from 2000 to 2019
#| echo: false

ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
  geom_point(alpha = 0.7, size = 2) + 
  labs(
    x = "Year",
    y = "Average Hourly Wage Rate (CAD)",
    color = "Education Level"
  ) +
  theme_minimal(base_size = 10) +  # Reduced base_size for smaller overall text
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),  # Smaller font for x-axis labels
    axis.text.y = element_text(size = 8),  # Smaller font for y-axis labels
    legend.position = "right",
    legend.text = element_text(size = 8)  # Smaller font for legend text
  ) +
  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
```
@fig-display-clean-data illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54. Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.

@@

The measurement of variables within this dataset was handled to ensure accuracy and relevance. The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income. This data is then vetted and standardized by statistical agencies to ensure it reflects accurate and fair representations of wages across different demographics.

@@

For an in-depth technical breakdown of the statistical analyses and additional visualizations, readers are directed to the appendices, where these aspects are discussed in detail. This ensures that the main text remains accessible while still providing resources for more technically inclined readers.

@@

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

$$
\begin{aligned}
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 +\beta_1x_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{aligned}
$$
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

The model is executed in R[@citeR] using the rstanarm package [@rstanarm]. Default priors from rstanarm are utilized, set with a mean of zero and a conservative standard deviation, reflecting no strong prior beliefs about the magnitude of the education coefficients.

@@

Given the economic theory and previous empirical findings, we anticipate that higher educational qualifications would correspond to higher hourly wages. This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$. Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing empirical evidence to support or refine this hypothesis.

Mathematical notation is incorporated using LaTeX syntax, enabling precise formulation of our statistical model. For example, the average effect of education on hourly wage rate is denoted by $\beta$.

@@

```{r}
#| label: tbl-summary-model
#| tbl-cap: Model Summary
#| echo: false
#| warning: false

first_model <- readRDS(file = here::here("models/first_model.rds"))

modelsummary(
  list(
    "Linear Model" = first_model
  )
)
```

@@

```{r fig.width=7, fig.height=4}
#| label: fig-visualize-model
#| fig-cap: Model Visualization
#| echo: false
#| warning: false

@@

In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate. This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities. The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion. In the next @sec-discussion, we will delve into the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.

@@

Our findings reinforce the perception of education as an investment yielding tangible financial returns. The analysis showcases a clear, positive trajectory in wage rates corresponding with ascending educational levels. This outcome underlines the vital role of education in driving economic growth, suggesting that educational enhancements could be pivotal in strategies aimed at combating poverty and fostering economic resilience.

Further examination reveals the potential of education as a tool for social mobility. The data indicates that as individuals climb the educational ladder, they unlock higher earnings, which could lead to a more equitable distribution of income. These insights could serve as a valuable asset for policymakers striving for educational equity and highlight the importance of making higher education accessible to all strata of society.

@@

The positive correlation observed in this study goes beyond confirming a long-suspected link between education and wages; it also points to the complex interplay of various wage determinants. Despite the clear trend, the dispersion of wage data around the regression line suggests multiple underlying factors at play. This finding suggests a need for a comprehensive approach to wage enhancement, integrating elements such as vocational training, relocation incentives, and specialized educational programs.

Expanding on the above, it is evident that education serves as one of many pillars supporting an individual's earning capacity. While education remains a significant factor, the disparity between the predicted and observed wages implies other variables are also influential. Recognizing this complexity is crucial for a complete understanding of how education impacts wage rates and the formulation of holistic policies that address the multifaceted nature of the job market.

## Limitations

The primary limitation of this study is its relatively small sample size, with only 180 observations. This constraint may impact the robustness and generalizability of the findings. Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers. Such a limited scope might overlook the nuanced effects of education on wages across different life stages.

## Future Research Directions

Future research should aim to disaggregate the age groups to understand the differential impact of education on wages across a more granular age spectrum. Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more nuanced insights into the education-wage dynamic at different life stages. Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.

Incorporating a more comprehensive range of variables and exploring more complex, perhaps non-linear models, would allow for a better understanding of the multifarious factors influencing wages. Such an enriched analysis could serve as the foundation for more effective, tailored economic policies that address both the direct and indirect ways education can influence economic outcomes.

@@

# Data Manipulation and Cleaning {#sec-appendix}

Most of the data in our dataset was previously cleaned for the project we are replicating. Thus, @fig-trends-in-birth-rate directly used data from `outputs/data/fig_1.csv` without any further cleaning necessary, as it simply displays the birth rate from 1980 to 2020. Thus, the cleaned data is directly stored to `outputs/data/fig_1.csv` from `inputs/data/fig_1.csv`.

In @fig-trends-in-birth-rate-in-young-people, the dataset was imported from `inputs/data/fig_2.csv` and used `dpylr` [@citedplyr] to select the birth rates of the six age groups (from 15 to 44). Furthermore, the selected data was renamed using `tidyverse` [@citetidyverse] and changed into names that represent the content of the data better. Moreover, these data is then pivoted using `tidyverse` [@citetidyverse] in order to be graphed properly using `ggplot2` [@citeggplot].

In @fig-birth-rate-comparison-old-and-young, the first dataset was imported from `inputs/data/fig_3.csv` and the dataset containing American states information was from `mapdata` [@citemapdata]. The first dataset was first sliced to remove birth rate information regarding the states of Alaska and Hawaii, as these two states can't be properly shown in the dataset from `mapdata` [@citemapdata]. Then, a temporary data frame is created with tibble from `tidyverse` [@citetidyverse] with the abbreviated state names from `inputs/data/fig_3.csv` and the state full names. Moreover, the abbreviated state name in the first dataset is switched with the state full name using left_join from `dpylr` [@citedplyr], which is then selected and renamed using `tidyverse` [@citetidyverse]. After this, the adjusted dataset with the state full names and birth rate is merged with the second dataset that contains the American states information for further graphing with `haven` [@citehaven].

In all three figures, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.

The transformation from raw to analysis-ready data involved substantial cleaning, including filtering by geography (Canada), type of work (both full- and part-time), and specific wage type (average hourly rate). Moreover, we excluded broad or ambiguous categories like 'Some high school' and 'Total, all education levels' to ensure clarity in our analysis. The resulting 'Education_numeric' variable was constructed using a sorted factorization process, mapping each education level to a corresponding numeric value.

In presenting our findings, we will incorporate graphs that visualize the relationship between education and wage rates, providing an intuitive understanding of the data's characteristics. Tables summarizing the observations, along with discussions on the graphs, will be included to elucidate key trends and observations. Summary statistics and any notable relationships between the variables will also be featured, offering a detailed overview of the dataset. Should the volume of this information warrant it additional detail, these elements may be relegated to appendices to maintain the accessibility of the data section.

>>>>

library(tidyverse)
library(dplyr)
library(here)
library(ggplot2)
library(haven)
library(kableExtra)
library(modelsummary)
library(rstanarm)
```

@@

The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. @sec-model then covers the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation. After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data. @sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.

## Estimand

In this study, our primary focus is to estimate the effect of education on hourly wages in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.

@@

This study uses a dataset sourced from the Open Government Portal of Canada [@citedata], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce. The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.

@@

The data was processed and cleaned using R[@citeR], a powerful statistical programming language. The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends. Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research. This step was crucial to maintaining the clarity and relevance of the analysis. For key operations, please refer to the [Appendix -@sec-data-cleaning].

@@

```{r}
#| label: tbl-display-clean-data
#| tbl-cap: First Ten Rows of Hourly Wage Across 2000-2019 After Data Processing
#| echo: false

@@

```{r fig.width=7, fig.height=3}
#| label: fig-display-clean-data
#| fig-cap: Average Hourly Wages by Educational Attainment, 2000-2019
#| echo: false

ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
  geom_point(alpha = 0.7, size = 2) + 
  labs(
    x = "Year",
    y = "Average Hourly Wage Rate (CAD)",
    color = "Education Level"
  ) +
  theme_minimal(base_size = 10) + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
```
@fig-display-clean-data built by `ggplot2` [@citeggplot] illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54. Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.

@@

The measurement of variables within this dataset was handled to ensure accuracy and relevance. The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income. This data is then vetted and standardized by the Government of Ontario to ensure it reflects accurate and fair representations of wages across different demographics.

@@

@@

In our analysis, we utilized a simple linear regression model, a technique well-suited for examining the relationship between a continuous dependent variable and one or more independent variables. Given that our dependent variable, `average hourly wage rate`, is continuous, a normal or Gaussian linear regression model is appropriate. This model assumes that the distribution of the wage rate, given the level of education, follows a normal distribution. This assumption allows for a straightforward interpretation of the parameters and is commonly used in practice. The use of a normal distribution for the error term is a standard choice, as it facilitates the estimation of the model parameters using maximum likelihood estimation. The simplicity and interpretability of this model, coupled with its appropriateness for the data at hand, make it a good fit for our study's objectives.

Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

$$
\begin{aligned}
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 +\beta_1x_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{aligned}
$$

The model is executed in R [@citeR] using the `rstanarm` package [@citerstanarm]. Default priors from `rstanarm` [@citerstanarm] are used, set with a mean of zero and a conservative standard deviation.

@@

Given the economic theory and previous findings, we anticipate that higher educational qualifications would correspond to higher hourly wages. This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$. Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing evidence to support or refine this hypothesis.

We selected a Gaussian linear regression model to examine the relationship between education level and average hourly wages because our dependent variable, the wage rate, is a continuous data point that typically follows a normal distribution. Linear regression is a straightforward method for assessing the impact of one or more independent variables on a dependent variable. In this case, it helps us understand how increasing education levels, which we have numerically encoded, relate to wage rates.

Further justification for this model comes from its foundation in the central limit theorem, which suggests that with a large sample size, our wage rate data should be normally distributed. Moreover, the expected positive correlation between education and wages aligns with existing economic studies, reinforcing our theoretical framework. The simplicity of the model aids in avoiding overfitting, ensuring that the results are generalizable and relevant to a broader population. The regression model's assumptions—linearity, independence, and normality of residuals—are all conditions that our dataset reasonably meets, making a Gaussian linear regression not only a convenient choice but also statistically appropriate for our analysis.

@@

```{r}
#| label: tbl-summary-model
#| tbl-cap: Summary of the Linear Regression Model
#| echo: false
#| warning: false

linear_model <- readRDS(file = here::here("models/linear_model.rds"))

modelsummary(
  list(
    "Linear Model" = linear_model
  )
)
```

@@

```{r fig.width=7, fig.height=4}
#| label: fig-visualize-model
#| fig-cap: Relationship between Hourly Wage Rate and Education Level
#| echo: false
#| warning: false

@@

In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate. This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities. The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion. In the next @sec-discussion, we will discuss the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.

@@

Our analysis reveals a clear, positive correlation between education level and average hourly wage rates across the data spanning from 2000 to 2019. The consistent upward trend in wage rates with higher education levels across the years underscores the value of educational attainment in the labor market. Despite some fluctuations, the general pattern suggests that investments in education are likely to be economically beneficial for individuals in the age group of 25-54 years.

@@

The results of this study offer clear evidence that educational attainment is closely tied to wage earnings, highlighting the premium that the labor market places on higher qualifications. This finding is consistent with current labor market theories that emphasize the importance of human capital—suggesting that investments in education are indeed reflected in higher wages. Specifically, our data supports the view that a well-educated workforce is critical for economic growth and competitiveness in today's globalized world, which increasingly values knowledge and specialized skills.

From a policy perspective, these insights could serve as a catalyst for reform, underscoring the need for investments in education. Policymakers might use this information to support sectors that are likely to offer greater economic returns for educational investments. This could involve bolstering STEM (Science, Technology, Engineering, and Mathematics) education, for example, or providing incentives for pursuing in-demand fields such as healthcare or information technology.

In conclusion, this study's findings highlight the significance of education as a determinant of economic success and reinforce the importance of informed educational and economic policymaking. For policymakers, the imperative is clear: to design and implement educational strategies that not only increase access to education but also ensure that the education provided is relevant and responsive to the current and future needs of the labor market.

## Limitations and Future Research Directions

The primary limitation of this study is its relatively small sample size, with only 180 observations. This constraint may impact the generalization of the findings. Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers. Such a limited scope might overlook the effects of education on wages across different life stages.

Future research should aim to disaggregate the age groups to understand the differential impact of education on wages. Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more insights into the education-wage dynamic at different life stages. Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.

@@

# Data Manipulation and Cleaning {#sec-data-cleaning}

During the data cleaning phase, the R packages `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], and `arrow` [@citearrow] were utilized. The raw data was imported using `read_csv` from the `tidyverse` [@citetidyverse] package. Subsequent operations filtered the dataset to focus on Canadian data, both full-time and part-time. Additionally, the focus was placed on records that reported the average hourly wage rate, while entries with ambiguous educational levels such as "Some high school" and aggregate categories like "Total, all education levels" were excluded to maintain data clarity and relevance.

The analysis was specifically targeted at the demographic group aged between 25 and 54 years, as this range represents a prime working-age population. Columns important to the analysis were chosen and renamed for better readability and straightforward reference in subsequent analytical procedures. This process included renaming the columns for the year and average hourly wage rate to 'Year' and 'Avg hourly wage rate' respectively.

Education levels were organized in a logical order ranging from "0 - 8 years" of education to "Above bachelor's degree" using the `factor` function to convert them into a categorical variable with a specified level order. This categorization was further enhanced by creating an accompanying numeric variable that mapped these ordered education levels to integers, thus facilitating quantitative analysis.

The final step involved saving the cleaned and structured data. This was done using the `write_csv` function to generate a CSV file for broad compatibility and the `write_parquet` function from the `arrow` [@citearrow] package for a more compressed and efficient file format, both of which were stored in the `data/analysis_data` directory.

In all figures and tables, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$). The density of the replicated data largely coincides with the observed data, suggesting that our model does a reasonable job of capturing the underlying distribution of the observed average hourly wage rates. This alignment is crucial as it indicates that the model's predictions are consistent with the real-world data we aim to understand and that the model can be a reliable tool for further inference.

In @fig-ppcheckandposteriorvsprior-2, we compare the posterior distributions of our parameters with the priors. This comparison reveals how the observed data have updated our beliefs about the parameters. We see that the posterior distributions for the intercept and the education_numeric coefficient are more concentrated and shifted from the priors, indicating that the data provided substantial information to update our initial assumptions. The sigma parameter's posterior, which represents the standard deviation of the residuals, is also refined from its prior, showing that the data have informed us about the variability in wages that is not explained by education level alone.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(linear_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(linear_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(linear_model, "trace")

plot(linear_model, "rhat")
```

In @fig-stanareyouokay-1, the trace plot demonstrates the sampling paths of the Markov chains for each parameter in the model. The trace for the intercept, education_numeric coefficient, and sigma parameter all display a ‘hairy caterpillar’ shape, indicating good mixing and convergence across multiple chains. This suggests that the model's parameter estimates are stable and reliable, as evidenced by the overlapping and intertwined paths, which is indicative of a well-fitting model.

@fig-stanareyouokay-2, the Rhat plot, presents the R-hat statistics for each parameter, which is a measure of convergence. We see that all values at or very close to 1.00. This is a desirable outcome, implying that further sampling would likely yield little additional information and that the posterior estimates can be considered reliable for inference.
